---
title: "Spatial DSEM MVP: Spatio-temporal SPDE Lognormal Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial DSEM MVP: Spatio-temporal SPDE Lognormal Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette demonstrates a minimal viable implementation of a spatial
density-surface evaluation model (DSEM) within FIMS. The model combines:

1. **Spatial SPDE prior**: A Matérn covariance field represented via SPDE
   finite-element matrices from `fmesher`.
2. **AR1 temporal evolution**: The latent spatial field evolves over years
   with first-order auto-regressive structure.
3. **Lognormal observation likelihood**: Positive survey observations are
   modelled as lognormal with a linear predictor on the log scale.

### Model specification

Let $\omega_{tk}$ denote the latent field at mesh node $k$ in year $t$
($t = 1, \ldots, T$; $k = 1, \ldots, n$).

**SPDE precision**: $\mathbf{Q} = \tau^2(\kappa^4 \mathbf{C}_0 +
2\kappa^2 \mathbf{C}_1 + \mathbf{C}_2)$

where $\mathbf{C}_0$, $\mathbf{C}_1$, $\mathbf{C}_2$ are FEM matrices
from the mesh.

**Initial condition** ($t = 0$): $\boldsymbol{\omega}_0 \sim \text{GMRF}(\mathbf{Q})$

**AR1 evolution** ($t > 0$): $\boldsymbol{\omega}_t - \rho \boldsymbol{\omega}_{t-1}
\sim \text{GMRF}(\mathbf{Q})$

**Observation likelihood**: For observation $i$ at location $\mathbf{s}_i$
in year $t_i$:

$$\log y_i \sim \mathcal{N}(\eta_i,\; \sigma_\text{obs}^2) - \log y_i$$

$$\eta_i = \beta_0 + (\mathbf{A} \boldsymbol{\omega}_{t_i})_i$$

where $\mathbf{A}$ is the bilinear projector from mesh nodes to observation
locations.

### Parameters

| Parameter     | Scale            | Description                               |
|---------------|------------------|-------------------------------------------|
| $\beta_0$     | log              | Intercept                                 |
| $\rho$        | logit            | AR1 temporal correlation ($\in (-1, 1)$) |
| $\kappa$      | log              | SPDE range parameter                      |
| $\tau$        | log              | SPDE variance parameter                   |
| $\sigma_\text{obs}$ | log        | Observation error standard deviation      |
| $\boldsymbol{\omega}$ | natural  | Latent field (random effects)             |

## Setup

```{r setup}
library(FIMS)
```

The following packages are required for this vignette:

```{r check-packages, eval=FALSE}
# Install fmesher if needed:
# install.packages("fmesher")
library(fmesher)
library(Matrix)
library(TMB)
```

```{r load-packages, include=FALSE}
has_fmesher <- requireNamespace("fmesher", quietly = TRUE)
has_matrix  <- requireNamespace("Matrix",  quietly = TRUE)
has_tmb     <- requireNamespace("TMB",     quietly = TRUE)
run_full    <- has_fmesher && has_matrix && has_tmb
```

## Build a simple mesh with `fmesher`

We create a small mesh over a 2D domain using a grid of known locations.

```{r build-mesh, eval=run_full}
library(fmesher)
library(Matrix)
library(TMB)

set.seed(42)

# Define a small set of known spatial locations on the unit square
n_loc <- 9
loc <- as.matrix(expand.grid(x = seq(0, 1, length.out = 3),
                              y = seq(0, 1, length.out = 3)))

# Build a mesh with fmesher (max.edge controls triangle size)
mesh <- fmesher::fm_mesh_2d(
  loc     = loc,
  max.edge = c(0.5, 1.5),   # inner / outer max edge lengths
  offset   = c(0.2, 0.5)
)

cat("Mesh nodes:", mesh$n, "\n")
```

## Build SPDE matrices and projector A

```{r spde-matrices, eval=run_full}
# Compute FEM matrices from the mesh
fem <- fmesher::fm_fem(mesh)
# fem$c0: mass matrix     (n x n)
# fem$g1: mixed           (n x n)
# fem$g2: stiffness       (n x n)

# Build projection matrix A: n_loc x n_nodes
# A[i, k] = basis-function weight for location i at node k
A_mat <- fmesher::fm_basis(mesh, loc = loc)

cat("C0 dimensions:", dim(fem$c0), "\n")
cat("A  dimensions:", dim(A_mat), "\n")
```

## Simulate toy observations

We simulate observations from the true model to have something to fit.

```{r simulate-data, eval=run_full}
n_years <- 3L
n_loc_per_year <- nrow(loc)
n_obs <- n_loc_per_year * n_years

# True parameters
beta0_true        <- 1.0
rho_true          <- 0.7
log_kappa_true    <- log(2.0)
log_tau_true      <- log(1.0)
log_sigma_obs_true <- log(0.2)

# Build Q from true kappa/tau
kappa_true <- exp(log_kappa_true)
tau_true   <- exp(log_tau_true)
C0 <- fem$c0
G1 <- fem$g1
G2 <- fem$g2
Q_true <- tau_true^2 * (kappa_true^4 * C0 + 2 * kappa_true^2 * G1 + G2)
Q_true <- (Q_true + Matrix::t(Q_true)) / 2 + Diagonal(nrow(Q_true)) * 1e-6

# Simulate spatial fields via Cholesky of precision
n_nodes_mesh <- mesh$n
R_chol <- Matrix::chol(Q_true)  # upper-triangular Cholesky of Q

# Simulate omega: AR1 over years
omega_sim <- matrix(0, nrow = n_years, ncol = n_nodes_mesh)
for (t in seq_len(n_years)) {
  z <- rnorm(n_nodes_mesh)
  # Solve R^T * R * omega = z  => omega = R^{-1} * R^{-T} * z
  omega_sim[t, ] <- as.numeric(Matrix::solve(R_chol, z))
  if (t > 1) {
    omega_sim[t, ] <- rho_true * omega_sim[t - 1, ] + omega_sim[t, ]
  }
}

# Generate observations
y_obs       <- numeric(n_obs)
obs_year_idx <- integer(n_obs)
for (t in seq_len(n_years)) {
  idx  <- ((t - 1) * n_loc_per_year + 1):(t * n_loc_per_year)
  eta  <- beta0_true + as.numeric(A_mat %*% omega_sim[t, ])
  y_obs[idx]        <- exp(rnorm(n_loc_per_year, mean = eta,
                                  sd = exp(log_sigma_obs_true)))
  obs_year_idx[idx] <- t - 1L  # 0-based year index
}

cat("Observations simulated:", n_obs, "\n")
cat("Min/max y_obs:", round(min(y_obs), 3), "/", round(max(y_obs), 3), "\n")
```

## Convert sparse matrices to FIMS triplets

```{r convert-triplets, eval=run_full}
C0_trips <- sparse_to_triplets(fem$c0)
C1_trips <- sparse_to_triplets(fem$g1)
C2_trips <- sparse_to_triplets(fem$g2)
A_trips  <- sparse_to_triplets(
  # Stack A for all observations (same spatial locations each year)
  do.call(rbind, replicate(n_years, A_mat, simplify = FALSE))
)

# Note: obs_year_idx already aligns with the stacked rows of A
cat("C0 non-zeros:", length(C0_trips$v), "\n")
cat("A  non-zeros:", length(A_trips$v), "\n")
```

## Configure the FIMS spatiotemporal SPDE model

```{r configure-fims, eval=run_full}
clear()  # Reset FIMS state

spde_obj <- create_spatiotemporal_spde(
  n_nodes           = mesh$n,
  n_years           = n_years,
  C0_triplets       = C0_trips,
  C1_triplets       = C1_trips,
  C2_triplets       = C2_trips,
  A_triplets        = A_trips,
  y_obs             = y_obs,
  obs_year_idx      = obs_year_idx,
  beta0_init        = 1.0,
  rho_logit_init    = 0.0,       # rho = tanh(0) = 0
  log_kappa_init    = 0.0,
  log_tau_init      = 0.0,
  log_sigma_obs_init = log(0.3),
  estimate_beta0    = TRUE,
  estimate_rho      = TRUE,
  estimate_log_kappa = FALSE,    # fix log_kappa
  estimate_log_tau   = FALSE,    # fix log_tau
  estimate_log_sigma_obs = TRUE
)

cat("SPDE object ID:", spde_obj$get_id(), "\n")
```

## Build the TMB model

```{r create-tmb, eval=run_full}
CreateTMBModel()

p  <- get_fixed()
re <- get_random()

cat("Fixed-effect parameters:", length(p), "\n")
cat("Random-effect parameters (omega):", length(re),
    "(=", mesh$n, "nodes x", n_years, "years)\n")
```

## Fit the model

```{r fit-model, eval=run_full}
obj <- TMB::MakeADFun(
  data       = list(),
  parameters = list(p = p, re = re),
  random     = "re",
  DLL        = "FIMS",
  silent     = TRUE
)

# Evaluate NLL at initial values
nll_init <- obj$fn(obj$par)
cat("NLL at initial values:", round(nll_init, 3), "\n")

# Optimise
opt <- nlminb(
  start     = obj$par,
  objective = obj$fn,
  gradient  = obj$gr,
  control   = list(iter.max = 300, eval.max = 500)
)
cat("Convergence:", opt$convergence,
    "(0 = success)\n")
cat("NLL at optimum:", round(opt$objective, 3), "\n")
```

## Extract parameter estimates

```{r estimates, eval=run_full}
# Update Information with final parameter values
set_fixed(opt$par)
set_random(rep(0, length(re)))  # RE marginalised by TMB

# SD report for uncertainty
sdr    <- TMB::sdreport(obj)
est    <- summary(sdr, "fixed")

cat("\nFixed-effect parameter estimates:\n")
print(round(est, 4))

# Extract estimated beta0 and rho
p_names <- get_parameter_names()
beta0_est     <- est[grep("beta0",     p_names), "Estimate"]
rho_logit_est <- est[grep("rho_logit", p_names), "Estimate"]
rho_est       <- tanh(rho_logit_est)

cat("\nEstimated beta0:", round(beta0_est, 3),
    "(true:", beta0_true, ")\n")
cat("Estimated rho:  ", round(rho_est, 3),
    "(true:", rho_true, ")\n")
```

## Visualise fitted spatial field (year 1)

```{r plot-field, eval=run_full, fig.width=6, fig.height=4}
# Get marginal estimates of omega
omega_est <- summary(sdr, "random")[, "Estimate"]
omega_mat <- matrix(omega_est, nrow = n_years, ncol = mesh$n, byrow = FALSE)

# Predicted log-density at observation locations for year 1
eta_pred_y1 <- beta0_est + as.numeric(A_mat %*% omega_mat[1, ])
plot(loc[, 1], loc[, 2],
     pch  = 21,
     bg   = colorRampPalette(c("blue", "white", "red"))(20)[
       cut(eta_pred_y1, 20)],
     cex  = 2,
     xlab = "x", ylab = "y",
     main = "Fitted log-density (Year 1)")
```

## Cleanup

```{r cleanup, eval=run_full}
clear()
```

## Summary

This vignette demonstrated a working MVP spatial DSEM in FIMS:

- **fmesher** was used to build the mesh and compute FEM matrices.
- **FIMS** estimated the fixed effects ($\beta_0$, $\rho$) and marginalised
  the random effects ($\boldsymbol{\omega}$) via TMB's Laplace approximation.
- The model recovers the simulated parameters reasonably well for a small
  mesh and few years.

For a real application, use a finer mesh (more nodes), more years of data,
and consider estimating $\kappa$ and $\tau$ jointly.
